%% ============================================================
%%  frontmatter/acknowledgement.tex
%% ============================================================
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\thispagestyle{plain}

%% Write your acknowledgements here.
I would like to express my sincere gratitude to my supervisor for the guidance, feedback, and encouragement provided throughout this research.

I am also grateful to Dr.~Kalpani Manathunga, Head of the Department of Software Engineering, Faculty of Computing, and Prof.~Nuwan Kodagoda, Deputy Vice-Chancellor -- Computing, Faculty of Computing, for their leadership and support in maintaining an academic environment that encourages applied research and innovation.

\medskip

I would like to extend my sincere appreciation to Prof.~Koliya Pulasinghe, Senior Professor, Department of Information Technology, Faculty of Computing, for his guidance and coordination as the Lecturer in Charge for the Research Project (Comprehensive Design and Analysis Project).

\medskip

I would like to thank the GradeLoop development team and the staff of the Department of Software Engineering at SLIIT for their technical assistance and collaborative support during the study.

\medskip

I sincerely thank the SLIIT students who volunteered for the evaluation by completing enrollment sessions and test coding assignments. Their participation made it possible to evaluate the BLAIM module using real user keystroke data.

\medskip

I would also like to thank my colleagues and peers who provided valuable insights and feedback during the development and testing of this module.

\medskip

Special thanks to my family and friends for their unwavering patience and moral support throughout this research journey.

\medskip

This research was carried out with support from the GradeLoop project infrastructure and the BLAIM evaluator environment used for implementation and testing.

\clearpage
